---
title: CordBat correction for metabolomics
author: 
- name: Nick Borcherding
  email: ncborch@gmail.com
  affiliation: Washington University in St. Louis, School of Medicine, St. Louis, MO, USA

date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'

output:
  BiocStyle::html_document:
    toc_float: true
package: CordBat
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Using CordBat}
  %\VignetteEncoding{UTF-8} 
---

# Load Libraries 

```{r}
library(CordBat)
library(ggplot2)
```

# Generate Synthetic Data

The synthetic dataset is engineered to mirror realistic mass spectrometry outputs by simulating 25 metabolites across six distinct batches (five experimental batches plus one reference batch). For each sample, baseline metabolite values are generated from a normal distribution with a high mean (e.g., 100000) to reflect typical intensity scales, with slight random noise. A batch effect is introduced by scaling a secondary normal distribution (with its own mean and minimal variability) and adding this offset to the baseline values. In addition, a grouping variable assigns each sample into one of two groups, enabling the assessment of both batch-related and intrinsic sample variability.

```{r}
# Set seed for reproducibility
set.seed(42)

# Define parameters
num_metabolites   <- 25         # number of metabolites
samples_per_batch <- 10         # samples per batch 
batches           <- c("Ref", "Batch1", "Batch2", "Batch3", "Batch4", "Batch5")

# Define batch effects for each batch
batch_effects <- list(
  "Ref"    = 0.0,
  "Batch1" = 1.0,
  "Batch2" = -1.0,
  "Batch3" = 0.5,
  "Batch4" = -0.5,
  "Batch5" = 1.5
)

# Create an empty list to store sample records
data_list <- list()
sample_counter <- 1

# Loop over each batch and each sample within the batch
for (batch in batches) {
  for (i in 1:samples_per_batch) {
    # Create a unique sample ID
    sample_id <- paste(batch, "Sample", i, sep = "_")
    
    # Assign Group: first half as "Group1", second half as "Group2"
    group <- if (i <= samples_per_batch / 2) "Group1" else "Group2"
    
    # Generate baseline metabolite values from a normal distribution 
    baseline <- rnorm(num_metabolites, mean = 100000, sd = 10000)
    
    # Simulate a realistic batch effect by multiplying with a random factor 
    met_values <- baseline + batch_effects[[batch]] * rnorm(num_metabolites, mean = 25000, sd = 5000)
    
    # Round the metabolite values to 2 decimal places
    met_values <- round(met_values, 2)
    
    # Name the metabolite variables (Metabolite1, Metabolite2, …)
    names(met_values) <- paste0("Metabolite", 1:num_metabolites)
    
    # Combine sample information with metabolite values.
    record <- c(SampleID = sample_id, Batch = batch, Group = group, met_values)
    
    # Store the record in the list
    data_list[[sample_counter]] <- record
    sample_counter <- sample_counter + 1
  }
}

# Convert the list of records into a data frame.
cordbat_example <- do.call(rbind, data_list)
cordbat_example <- as.data.frame(cordbat_example, stringsAsFactors = FALSE)

# Convert metabolite columns from character to numeric
metabolite_cols <- grep("Metabolite", names(cordbat_example))
cordbat_example[ , metabolite_cols] <- lapply(cordbat_example[ , metabolite_cols], as.numeric)

# Optionally, convert Batch and Group to factors
cordbat_example$Batch <- factor(cordbat_example$Batch, levels = batches)
cordbat_example$Group <- factor(cordbat_example$Group, levels = c("Group1", "Group2"))
```

## Visualizing Batch/Group with PCA

```{r}
# Identify the columns with metabolite measurements
metabolite_cols <- grep("Metabolite", names(cordbat_example))

# Perform PCA on the metabolite data; scaling is recommended
pca_res <- prcomp(cordbat_example[, metabolite_cols], scale. = TRUE)

# Combine PCA results with the metadata
pca_df <- data.frame(
  SampleID = cordbat_example$SampleID,
  Batch    = cordbat_example$Batch,
  Group    = cordbat_example$Group,
  PC1      = pca_res$x[, 1],
  PC2      = pca_res$x[, 2]
)

# Plot PCA colored by Batch
p_batch <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Batch)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA of Mass Spec Data - Colored by Batch",
       x = "Principal Component 1",
       y = "Principal Component 2")

# Plot PCA colored by Group
p_group <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA of Mass Spec Data - Colored by Group",
       x = "Principal Component 1",
       y = "Principal Component 2")

print(p_batch)
print(p_group)
```


