---
title: CordBat correction for metabolomics
author: 
- name: Nick Borcherding
  email: ncborch@gmail.com
  affiliation: Washington University in St. Louis, School of Medicine, St. Louis, MO, USA

date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'

output:
  BiocStyle::html_document:
    toc_float: true
package: CordBat
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Using CordBat}
  %\VignetteEncoding{UTF-8} 
---

# Load Libraries 

```{r}
library(CordBat)
library(ggplot2)
```

# Generate Synthetic Data

The synthetic dataset is engineered to mirror realistic mass spectrometry outputs by simulating 25 metabolites across six distinct batches (five experimental batches plus one reference batch). Unlike a uniform baseline, each metabolite is assigned a unique baseline intensity drawn from a normal distribution with a high mean (around 100,000) and moderate variability (e.g., SD ≈ 10,000) to better reflect inherent differences in metabolite abundance. For each sample, small random noise is added to these metabolite-specific values, mimicking the natural variability seen in mass spec measurements. A batch effect is then introduced by scaling a secondary normal distribution, with its own mean (approximately 25,000) and minimal variability, by a batch-specific coefficient and adding the resulting offset to the baseline values. This simulates systematic technical deviations between batches.

In addition to the batch effects, a grouping variable assigns each sample into one of two groups. To model biologically meaningful differences, samples in one group receive an additional fixed offset for a subset of metabolites (e.g., the first 10 metabolites). This deliberate group effect enables the assessment of both technical batch-related biases and intrinsic sample variations, thereby providing a rigorous framework for evaluating batch correction methods and community detection algorithms.

```{r}
# Set seed for reproducibility
set.seed(42)

# Define parameters
num_metabolites   <- 25         # number of metabolites
samples_per_batch <- 10         # samples per batch (should be even for balanced grouping)
batches           <- c("Ref", "Batch1", "Batch2", "Batch3", "Batch4", "Batch5")

# Define metabolite-specific baseline means to simulate realistic differences.
# Each metabolite’s mean is drawn from a normal distribution with mean = 100000 and sd = 5000.
met_means <- rnorm(num_metabolites, mean = 100000, sd = 10000)

# Define batch effects for each batch
batch_effects <- list(
  "Ref"    = 0.0,
  "Batch1" = 1.0,
  "Batch2" = -1.0,
  "Batch3" = 0.5,
  "Batch4" = -0.5,
  "Batch5" = 1.5
)

# Define a systematic group effect.
group_offset <- c(rep(500, 10), rep(0, num_metabolites - 10))

# Create an empty list to store sample records
data_list <- list()
sample_counter <- 1

# Loop over each batch and each sample within the batch
for (batch in batches) {
  for (i in 1:samples_per_batch) {
    # Create a unique sample ID
    sample_id <- paste(batch, "Sample", i, sep = "_")
    
    # Assign Group: first half as "Group1", second half as "Group2"
    group <- if (i <= samples_per_batch / 2) "Group1" else "Group2"
    
    # Generate baseline metabolite values with metabolite-specific means
    baseline <- rnorm(num_metabolites, mean = met_means, sd = 2)
    
    # Simulate a realistic batch effect by scaling a random factor 
    batch_effect_value <- batch_effects[[batch]] * rnorm(num_metabolites, mean = 25000, sd = 5000)
    
    # Compute the adjusted metabolite values as baseline plus batch effect
    met_values <- baseline + batch_effect_value
    
    # Apply group effect if sample is in Group2 
    if (group == "Group2") {
      met_values <- met_values + group_offset
    }
    
    # Round the metabolite values to 2 decimal places
    met_values <- round(met_values, 2)
    
    # Name the metabolite variables 
    names(met_values) <- paste0("Metabolite", 1:num_metabolites)
    
    # Combine sample information with metabolite values into a record
    record <- c(SampleID = sample_id, Batch = batch, Group = group, met_values)
    
    # Store the record in the list
    data_list[[sample_counter]] <- record
    sample_counter <- sample_counter + 1
  }
}

# Convert the list of records into a data frame.
cordbat_example <- do.call(rbind, data_list)
cordbat_example <- as.data.frame(cordbat_example, stringsAsFactors = FALSE)

# Convert metabolite columns from character to numeric
metabolite_cols <- grep("Metabolite", names(cordbat_example))
cordbat_example[, metabolite_cols] <- lapply(cordbat_example[, metabolite_cols], as.numeric)

# Optionally, convert Batch and Group columns to factors
cordbat_example$Batch <- factor(cordbat_example$Batch, levels = batches)
cordbat_example$Group <- factor(cordbat_example$Group, levels = c("Group1", "Group2"))
```

## Visualizing Batch/Group with PCA

```{r}
# Identify the columns with metabolite measurements
metabolite_cols <- grep("Metabolite", names(cordbat_example))

# Perform PCA on the metabolite data; scaling is recommended
pca_res <- prcomp(cordbat_example[, metabolite_cols], scale. = TRUE)

# Combine PCA results with the metadata
pca_df <- data.frame(
  SampleID = cordbat_example$SampleID,
  Batch    = cordbat_example$Batch,
  Group    = cordbat_example$Group,
  PC1      = pca_res$x[, 1],
  PC2      = pca_res$x[, 2]
)

# Plot PCA colored by Batch
p_batch <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Batch)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA of Mass Spec Data - Colored by Batch",
       x = "Principal Component 1",
       y = "Principal Component 2")

# Plot PCA colored by Group
p_group <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA of Mass Spec Data - Colored by Group",
       x = "Principal Component 1",
       y = "Principal Component 2")

print(p_batch)
print(p_group)
```


